### 2019-06-21 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
Proof reading Ch4

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.70\linewidth]{system_model1.eps}
    \caption[In our system model, some dynamic devices (in the IoT network) transmit packets to a gateway and suffer from the interference generated by neighboring networks.]{In our system model, some dynamic devices (in the \textcolor{blue}{IoT network in blue}) transmit packets to a gateway and suffer from the interference generated by neighboring networks (in \textcolor{orange}{orange} left/right).}
    \label{fig:41:system_model1}
\end{figure}

@@

\paragraph{Slotted protocol.}
%
For simplicity, and for consistency with the rest of this thesis, we only consider a discrete protocol (\ie, slotted).
As illustrated in Figure~\ref{fig:41:protocol}, we suppose a slotted protocol, in both time and frequency.
All devices share a synchronized time, and know in advance $K$, the finite number of available RF channels.
In each time slot, the devices try to send packets to the unique gateway, which listens continuously to all channels, following an ALOHA-based communication, with no sensing.
Each time slot is divided in two parts: first for uplink communications in which data packets are sent by end-devices to the gateway.
In one channel, if only one packet is sent in this part of the slot, the gateway can decode it and sends an acknowledgement to the device in the second part (on the same channel).
If two or more devices send an uplink packet in the same slot, the uplink packets collide, and the acknowledgement (\Ack) is not transmitted. In this case, we say that there is a \emph{collision} in this time slot.
In other words, if the gateway cannot successfully decode the incoming (uplink) messages because they were corrupted by collisions, it does not send any \Ack{} back.
This way, no collision can occur on the downlink messages, easing the analysis of collisions.

\begin{figure}[!h]
    \centering
    \includegraphics[scale=0.40]{protocol.eps}
    \caption[The considered time-frequency slotted protocol. Each frame is composed by a fixed duration uplink slot in which the end-devices transmit their (uplink) packets. If a packet is well received, the gateway replies by transmitting an \Ack, after the ack delay.]{The considered time-frequency slotted protocol. Each frame is composed by a fixed duration \textcolor{blue}{uplink slot} in which the end-devices transmit their \textcolor{blue}{(uplink) packets}. If a packet is well received, the gateway replies by transmitting an \textcolor{darkgreen}{\Ack}, after the ack delay.}
    \label{fig:41:protocol}
\end{figure}

@@

\textbf{Random emission patterns:}
%
We suppose that all devices follow the same emission pattern, being fixed in time, and we choose to model it as a simple Bernoulli process:
all devices have the same probability to send a packet in any (discrete) temporal slot, and we denote $p \in (0, 1)$ this probability\footnote{~In the experiments below, $p$ is about $10^{-3}$, because in a crowded network $p$ should be smaller than $K / (S + D)$ for all devices to communicate successfully (in average).}.
The parameter $p$ essentially controls the frequency of communication for each device, once the time scale is fixed, and $1/p$ is proportional to the \emph{duty cycle}.
% (\ie, real time during two messages)
For instance, for IoT devices sending messages with a duration of $1$ second and possibly at every second, then on a daily basis we have $p = 1 / (12 \times 60 \times 60) \simeq 1.5 \times 10^{-5}$.

\textbf{Some assumptions on the network occupation:}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the \SI{868}{\mega\hertz} ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for realistic IoT networks.\\
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
\indent
On the opposite, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to \emph{have an average of active devices not larger\footnote{~In Chapter~\ref{chapter:5} we simply consider the case of $p=1$ and $M = D \leq K$ dynamic devices, referred to as \emph{players}.} than the number of channels}.

\textbf{Link with the MAB model:}
%
All devices follow a Bernoulli emission process.
Consider the network from the point of view of one dynamic device:
every time a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $A(t) = k \in[K]$.
Then, the dynamic device waits in this channel $A(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r(t) \eqdef Y_{k,t} \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$\sum_{t = 1}^T r(t),$
which is the usual objective in MAB problems (see Section~\ref{sec:2:notations}).

@@

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $k$, at that time slot.
the $S_k$ static devices in each channel $k$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$,
so probability of successful transmission is computed as follows

@@

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $k$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device}\;k\;\text{did not sent in}\;i) = \prod_{k=1}^{D-1} \Pr(\text{device}\;k\;\text{did not sent in}\;i)$. And $\Pr(\text{device}\;k\;\text{sent in}\;i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_k}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{k=1}^{K} (1-p)^{S_k} .
\end{align}
%
This expression \eqref{eq:41:strategynaive} is constant (in time), and easy to compute numerically, but comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it
(maybe after a long enough initial learning period).

@@

Consequently, an optimal allocation vector $(D_1,\dots,D_{K})$ is a solution of the following real-valued constraint optimization problem:
%
\begin{subequations}\label{eq:41:prob}
\begin{align}
    % \begin{cases}
    \underset{D_1,\dots,D_{K}}{\arg\max}\; & \sum_{k=1}^{K} D_k (1 - p)^{S_k + D_k -1}, \label{eq:41:optPb}\\
    \text{such that}\;\; & \sum_{k=1}^{K} D_k = D, \label{eq:41:eqCstr}\\
    & D_k \geq 0 \qquad \forall k\in\llbracket 1;K\rrbracket . \label{eq:41:ineqCstr}
    % \end{cases}
\end{align}
\end{subequations}

@@

Where in the equation~\eqref{eq:41:Dilambda}, $(a)^{+} \eqdef \max(a,0)$, and $\mathcal{W}$ denotes the $\mathcal{W}$-Lambert function which is the reciprocal bijection of $x \mapsto x e^x$ on $\mathbb{R^+} = [0, +\infty)$ (which can be computed numerically in an efficient manner, \cite{Corless96}).
Moreover, condition \eqref{eq:41:eqCstr} implies that the Lagrange multiplier $\lambda$ is the solution of the constraint $\sum_{k=1}^{K} D_k^*(\lambda) = D$.
%
This single constraint can be solved numerically, with simple one-dimensional root finding algorithms.
Solving the optimization problem provides the optimal real number value for $D_k^*$, which has to be rounded\footnote{~Any rounding choice will give about the same repartition, up-to a difference of only one device by channel, and so we chose to round from below for the first channels.} to find the optimal number of devices for channel $k$:
%
$\widehat{D_k} = \lfloor D_k^* \rfloor$ for $1 \leq k < K$, and $\widehat{D_{K}} = D - \sum_{k=1}^{K - 1} \widehat{D_k}$.

@@

\textbf{Using a bandit algorithm for IoT devices:}
%
Our IoT application is challenging in that there are \emph{multiple} players (the dynamic devices) interacting with the \emph{same} arms (the channels), without any centralized communication (they do not even know the total number of dynamic devices).
%
% Considered alone, each dynamic device
We propose algorithms in which each dynamic device ignores all the other one and
implements a learning algorithm to play a bandit game.
%
In each time slot, if it has to communicate (which happens with probability $p$), then it chooses a channel and it receives a reward $1$ if the transmission is successful, $0$ otherwise.
Each device aims at maximizing the sum of the rewards collected during its communication instants, which shall indeed maximize the fraction of successful transmissions. Besides the modified time scale (rewards are no longer collected at every time step), this looks like a usual bandit problem.
However, it cannot be modeled as a stochastic MAB, as the rewards are (unfortunately) \emph{not} \iid: they not only depend on the (stationary, \iid) behavior of the static devices, but also on the behavior of other dynamic devices, that is not stationary (because of learning and random activation of each device).
%
Despite this, we show in the next subsection that running a stochastic bandit algorithm for each device based on its own rewards is surprisingly successful.

@@

Clearly, the problem is less interesting for a homogeneous repartition, as all channels appear the same for dynamic devices, and so even with $D$ small in comparison to $S$, the system behaves like in Figure~\ref{fig:41:100intelligent}, where the performance of the five approaches are very close.

@@

\textbf{Successful transmission rate as a function of the number of dynamic devices.}
%
To better assess the evolution of the optimal policy compared to the random one, we have displayed on Figure~\ref{fig:41:perf_learning} the evolution of the gain, in terms of successful transmission rate, provided by the optimal oracle and the two learning policies, after $10^6$ time slots, \ie, about $1000$ transmissions for each IoT device.
We can see that when the proportion of end-devices is low (\eg, $1\%$ of devices are dynamic), the optimal policy provides an improvement of $16\%$ compared to random channel selection.
The TS algorithm always provides near-optimal performance, but the \UCB{} algorithm has a lowest rate of convergence and performs consequently worse after $1000$ transmissions, for instance it only provides a gain of $12\%$ for the same proportion of dynamic devices ($1\%$),
for the considered values of $S_1,\dots,S_K$.

>>>>

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.70\linewidth]{system_model1.eps}
    \caption[In our system model, some dynamic devices (in the IoT network) transmit packets to a gateway and suffer from the interference generated by neighboring networks.]{In our system model, some dynamic devices (in the \textcolor{blue}{IoT network in blue}) transmit packets to a gateway and suffer from the interference generated by neighboring networks (in \textcolor{orange}{orange} left/right).}
    \label{fig:41:system_model1}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[scale=0.40]{protocol.eps}
    \caption[The considered time-frequency slotted protocol. Each frame is composed by a fixed duration uplink slot in which the end-devices transmit their (uplink) packets. If a packet is well received, the gateway replies by transmitting an \Ack, after the ack delay.]{The considered time-frequency ALOHA-based communication protocol \cite{Abramson1970,Roberts75}. Each frame is composed by a fixed duration \textcolor{blue}{uplink slot} in which the end-devices transmit their \textcolor{blue}{(uplink) packets}. If a packet is well received, the gateway replies by transmitting an \textcolor{darkgreen}{\Ack}, after the ack delay.}
    \label{fig:41:protocol}
\end{figure}

\paragraph{Slotted protocol.}
%
For simplicity, and as in all this thesis, we only consider a discrete protocol (\ie, slotted).
As illustrated in Figure~\ref{fig:41:protocol}, we suppose a slotted protocol, in both time and frequency.
All devices share a synchronized time, and know in advance $K$, the finite number of available RF channels.
In each time slot, the devices try to send packets to the unique gateway, which listens continuously to all channels, following an ALOHA-based communication protocol \cite{Abramson1970,Roberts75}, with no sensing.
Each time slot is divided in two parts: first for uplink communications in which data packets are sent by end-devices to the gateway.
In one channel, if only one packet is sent in this part of the slot, the gateway can decode it and sends an acknowledgement to the device in the second part (on the same channel).
If two or more devices send an uplink packet in the same slot, the uplink packets collide, and the acknowledgement (\Ack) is not transmitted. In this case, we say that there is a \emph{collision} in this time slot.
In other words, if the gateway cannot successfully decode the incoming (uplink) messages because they were corrupted by collisions, it does not send any \Ack{} back.
This way, no collision can occur on the downlink messages, easing the analysis of collisions.

@@

\textbf{Random emission patterns.}
%
We suppose that all devices follow the same emission pattern, being fixed in time, and we choose to model it as a simple Bernoulli process:
all devices have the same probability to send a packet in any (discrete) temporal slot, and we denote $p \in (0, 1)$ this probability.
The parameter $p$ essentially controls the frequency of communication for each device, once the time scale is fixed, and $1/p$ is proportional to the \emph{duty cycle}.
% (\ie, real time during two messages)
For instance, for IoT devices sending messages with a duration of $1$ second and possibly at every second, then on a daily basis we have $p = 1 / (12 \times 60 \times 60) \simeq 1.5 \times 10^{-5}$.
In the experiments below, $p$ is about $10^{-3}$, because in a crowded network $p$ should be smaller than $K / (S + D)$ for all devices to communicate successfully (in average).

\textbf{Some assumptions on the network occupation.}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the \SI{868}{\mega\hertz} ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for realistic IoT networks.\\
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
\indent
On the opposite, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to \emph{have an average of active devices not larger\footnote{~In Chapter~\ref{chapter:5} we simply consider the case of $p=1$ and $M = D \leq K$ dynamic devices, referred to as \emph{players}.} than the number of channels}.

\textbf{Link with the MAB model.}
%
All devices follow a Bernoulli emission process.
Consider the network from the point of view of one dynamic device:
every time a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $A(t) = k \in[K]$.
Then, the dynamic device waits in this channel $A(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r(t) \eqdef Y_{k,t} \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$\sum_{t = 1}^T r(t),$
which is the usual objective in MAB problems (see Section~\ref{sec:2:notations}).

@@

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $k$, at that time slot.
the $S_k$ static devices in each channel $k$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$,
so probability of successful transmission is computed as follows.

@@

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $k$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device } k \text{ did not sent in } i) = \prod_{k=1}^{D-1} \Pr(\text{device } k \text{ did not sent in } i)$. And $\Pr(\text{device } k \text{ sent in } i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_k}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{k=1}^{K} (1-p)^{S_k} .
\end{align}
%
This probability \eqref{eq:41:strategynaive} is constant wrt time, and easy to compute numerically.
Comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it,
maybe after a long enough initial learning period.

@@

Consequently, an optimal allocation vector $(D_1,\dots,D_{K}) \in \R^K$
is a solution of the following real-valued constraint optimization problem:
%
\begin{subequations}\label{eq:41:prob}
\begin{align}
    % \begin{cases}
    \underset{D_1,\dots,D_{K}}{\arg\max}\; & \sum_{k=1}^{K} D_k (1 - p)^{S_k + D_k -1}, \label{eq:41:optPb}\\
    \text{such that}\;\; & \sum_{k=1}^{K} D_k = D, \label{eq:41:eqCstr}\\
    & D_k \geq 0 \qquad \forall k\in [K]. \label{eq:41:ineqCstr}
    % \end{cases}
\end{align}
\end{subequations}

@@

Where in the equation~\eqref{eq:41:Dilambda}, $(a)^{+} \eqdef \max(a,0)$, and $\mathcal{W}$ denotes the $\mathcal{W}$-Lambert function which is the reciprocal bijection of $x \mapsto x e^x$ on $\mathbb{R^+} = [0, +\infty)$ (which can be computed numerically in an efficient manner, \cite{Corless96}).
Moreover, condition \eqref{eq:41:eqCstr} implies that the Lagrange multiplier $\lambda$ is the solution of the constraint $\sum_{k=1}^{K} D_k^*(\lambda) = D$.
This single constraint can be solved numerically, with simple one-dimensional root finding algorithms.

Note that solving the optimization problem provides the optimal \emph{real values} $(D_k^*)_k$, which have to be rounded to find the optimal \emph{integer} number of devices for channel $k$.
Any rounding choice will give about the same repartition, up-to a difference of only one device by channel, and so we chose to round from below for the first channels:
%
$\widehat{D_k} = \lfloor D_k^* \rfloor$ for $1 \leq k < K$, and $\widehat{D_{K}} = D - \sum_{k=1}^{K - 1} \widehat{D_k}$.

@@

\textbf{Using a bandit algorithm for IoT devices.}
%
Our IoT application is challenging in that there are \emph{multiple} players (the dynamic devices) interacting with the \emph{same} arms (the channels), without any centralized communication (they do not even know the total number of dynamic devices).
%
% Considered alone, each dynamic device
We propose algorithms in which each dynamic device ignores all the other one and
implements a learning algorithm to play a bandit game.
%
In each time slot, if it has to communicate (which happens with probability $p$), then it chooses a channel and it receives a reward $1$ if the transmission is successful, $0$ otherwise.
Each device aims at maximizing the sum of the rewards collected during its communication instants, which shall indeed maximize the fraction of successful transmissions. Besides the modified time scale (rewards are no longer collected at every time step), this looks like a usual bandit problem.
However, it cannot be modeled as a stochastic MAB, as the rewards are (unfortunately) \emph{not} \iid: they not only depend on the (stationary, \iid) behavior of the static devices, but also on the behavior of other dynamic devices, that is not stationary (because of learning and random activation of each device).
%
Despite this, we show in the next subsection that running a stochastic bandit algorithm for each device based on its own rewards is surprisingly successful.

@@

Clearly, the problem is less interesting for a homogeneous repartition, as all channels are equivalent for dynamic devices, and so even with $D$ small in comparison to $S$, the system behaves like in Figure~\ref{fig:41:100intelligent}, where the performance of the five approaches are very close.

@@

\paragraph{Successful transmission rate as a function of the number of dynamic devices.}
%
To better assess the evolution of the optimal policy compared to the random one, we have displayed on Figure~\ref{fig:41:perf_learning} the evolution of the gain, in terms of successful transmission rate, provided by the optimal oracle and the two learning policies, after $10^6$ time slots, \ie, about $1000$ transmissions for each IoT device.
We can see that when the proportion of end-devices is low (\eg, $1\%$ of devices are dynamic), the optimal policy provides an improvement of $16\%$ compared to random channel selection.
The TS algorithm always provides near-optimal performance, but the \UCB{} algorithm has a lowest rate of convergence and performs consequently worse after $1000$ transmissions, for instance it only provides a gain of $12\%$ for the same proportion of dynamic devices ($1\%$),
for the considered values of $S_1,\dots,S_K$.
